---
title: The Colorado Preliminary Congressional Map is Fair
description: |
  Colorado's Congressional Independent Redistricting Commission has released a 
  preliminary map for the 2022-2030 congressional districts.
author:
  - name: Christopher T. Kenny
    affiliation: Department of Government, Harvard University
    url: https://www.christophertkenny.com/
  - name: Cory McCartan
    affiliation: Department of Statistics, Harvard University
    url: https://corymccartan.github.io/
date: 06-26-2021
output:
  distill::distill_article:
    self_contained: false
bibliography: biblio.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)

library(here)
devtools::load_all(here("."))

analysis = run_analysis("CO", "cd", "prelim")
map = analysis$map
plans = analysis$results
```

```{r cd-map, fig.cap="The preliminary congressional districting plan.", fig.height=5}
plot_cds(map, get_existing(map), COUNTYFP20, "CO")
```

On June 23, Colorado's brand-new independent redistricting commission released a 
[preliminary redistricting plan](https://redistricting.colorado.gov/content/prelim-congressional-maps)
for the state's congressional districts.  Due to its above-average population 
growth, Colorado has gained an additional congressional seat since the last
time districts were drawn, in 2011.

Already, people on both sides of the aisle 
[are less-than-thrilled](https://www.cpr.org/2021/06/23/a-congressional-district-for-the-north-denver-metro-thats-where-the-first-map-has-it/)
with the proposed preliminary plans, citing concerns with district
competitiveness. Some have suggested that the plan is slightly
[biased to favor Republicans](https://www.politico.com/news/2021/06/23/redistricting-colorado-map-perlmutter-495830),
potentially making it easier for them to gain seats and recapture the House.

```{r co-map, layout="l-page", fig.width=12, fig.height=5, fig.cap="Partisan (2016 election) and demographic patterns in Colorado, with proposed districts overlaid."}
plot_partisan(map, dem_16, rep_16) + plot_minority(map, e_pop_white/e_pop)
```

But what constitutes a "fair" map in Colorado depends a lot how different
types of voters are distributed, where cities are, and more.  What appears at 
first to be a slightly GOP-friendly plan could in fact be a plan that is good
for Democrats, relative to what might otherwise be expected. In other words,
to talk about the Colorado plan, it's important to consider what the
alternatives might be.

We'll evaluate the Colorado plan by comparing it to 3,000 hypothetical other
plans, which were drawn by a computer algorithm following the same rules that 
the independent commission has to follow. 

# Generating Plans
Generating 3,000 hypothetical plans is not straightforward---there are vastly
more ways to draw redistricting plans than there are atoms in the universe, and
yet only a minuscule fraction of these plans satisfy all the relevant legal
constraints.
Here, we'll use a new algorithm [@mcca:imai:20], which randomly generates plans
according to a specific statistical distribution. Reflecting
[Colorado's redistricting criteria](https://redistricting.colorado.gov/content/congressional-redistricting),
we ensure the generated plans:

* Have districts with equal populations (at most 0.1% away from full equality).
* Have geographically contiguous and compact districts.
* Follow county lines as much as possible.
* Don't use any information on incumbents.

<aside>
In line with the commission's findings, we do not require the creation of a 
majority-minority district for compliance with the federal Voting Rights Act.
</aside>

Here are a few of the sampled plans: 

```{r hyp-map, layout="l-page", fig.width=12, fig.height=10}
plot_cds(map, as.matrix(plans)[,2], COUNTYFP20, "CO") +
    plot_cds(map, as.matrix(plans)[,3], COUNTYFP20, "CO") +
    plot_cds(map, as.matrix(plans)[,4], COUNTYFP20, "CO") +
    plot_cds(map, as.matrix(plans)[,5], COUNTYFP20, "CO") 
```

# Analysis
Once we've generated the set of comparison plans, we can use past election
data to estimate how congressional elections might have turned out under 
each of these alternative plans.  The results of this calculation are plotted
in Figure \@ref(fig:dem-distrs) below, for the 2016 and 2018 elections. Each dot
represents a district in a possible plan, with blue dots representing
Democratic-controlled districts and red dots Republican-controlled districts,
and the districts are arranged by how Democratic they are.  The black lines show
the estimated election results under the newly-proposed plan. When these lines
are lower than their corresponding set of points, that indicates that the
proposed district has fewer Democratic voters than would be expected under an
average fair and neutral map.

```{r dem-distrs, layout="l-body-outset", fig.width=9, fig.height=9, fig.cap="Democratic vote by district for the proposed plan and 3,000 alternative plans."}
dem_distr_plot(plans)
```

What we see overall is a proposed set of districts that is very much in line 
with what would be expected under a fair and neutral procedure.  The only
consistent difference is a lower number of Democratic voters in two
Republican-leaning districts, which is offset by slightly higher number of
Democratic voters in the most competitive district.  This feature of the plan
helps Democrats versus the counterfactual, which would lead to more reliable
4-4 splits.

***

# Data
Since the Commission provides population estimates at the 2010 block level, we
start by building a block dataset using `geomander` [@geomander]. This has the
shapes for each block along with the population and voting age populations by
block.

Generally, we use a different definition of populations from what Colorado
provides. We often treat race and ethnicity as one variable. However, Colorado's
preliminary data follows the Census Bureau's choice of treating race separately
from ethnicity. This means that Hispanic is separated from the standard race
categories. Additionally, the preliminary estimates do not include voting age
population. As such, we build additional data using the `blockpop` package
[@blockpop], which uses the [FCC's yearly block estimates](https://www.fcc.gov/staff-block-estimates) 
of populations to estimate the 2020 population at the block level. This is
fairly consistent and provides cleaner categories for 2020, but differs slightly
from Colorado's choice of population estimates.

As for election data, we rely on the precinct data from the [Voting and Election
Science Team](https://twitter.com/vest_team?lang=en), who produce some of the
best redistricting data available. We average the Republican votes and
Democratic votes across offices for 2016 by precinct across president and senate
races [@vest:16]. We do the same across the elections for Governor, Attorney
General, and Secretary of State elections in 2018 [@vest:18]. To match the
demographic data, we project this data down to blocks, splitting it by the
proportion of voting age population in each precinct.

With all of this data estimated at the 2010 Census block level, we aggregate to
2020 voting districts. Using the 2020 Census's shapefiles, we can then
approximate the redistricting process in Colorado. This gives us 3108 geographic
units, which average 1,618 people per voting district under the 2010 data, 1860
people under the data from blockpop, and 1858 under the preliminary data from
Colorado.

Finally, Colorado provides the district shapefiles, but doesn't provide a block
assignment file (BAF), which would match blocks to districts directly. Without
this, we match blocks to districts by the area overlap of each voting district
with the districts. Matching by area is imperfect, but this gives us a good
starting point to consider the preliminary plan on 2020 geographies.
