---
title: "Colorado Update: The Staff Plan is Worse"
description: |
    Staff for the Colorado Independent Congressional Redistricting Commission have released an updated congressional map.
author:
  - name: Cory McCartan
    affiliation: Department of Statistics, Harvard University
    url: https://corymccartan.github.io/
  - name: Christopher T. Kenny
    affiliation: Department of Government, Harvard University
    url: https://www.christophertkenny.com/
date: 09-07-2021
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, dev="ragg_png")
knitr::opts_knit$set(eval.after="fig.cap")

library(here)
devtools::load_all(here("."))

analysis = run_analysis("CO", "cd", "staff")
map = analysis$map
plans = analysis$results
```

<aside>
```{r results="asis"}
make_grade(plans$tol_01) %>%
    mutate(plan = c("Preliminary", "Staff")) %>%
    output_grades()
```
</aside>

This post follows up on a [previous analysis](../2021-06-25-colorado-preliminary-congressional-districts/) of Colorado's congressional redistricting plan.
Since then, staff for the Colorado Independent Congressional Redistricting Commission have released an updated congressional map.
This new map has already [been criticized](https://twitter.com/Redistrict/status/1435274464934211584) for its partisan balance and boundary lines.

```{r, results = 'hide'}
p1 = plot_cds(map, map$cd, county, "CO") + labs(title="Staff Plan")
p2 = plot_cds(map, map$cd_prelim, county, "CO") + labs(title="Preliminary Plan")
```

```{r cd-map, fig.width=9, fig.height=4, fig.cap="The preliminary and staff congressional districting plans."}
p1 + p2
```

# Political Features of Both Maps

[As with all of our analyses](../../methods.html), we compare these proposed plans with a large set of simulated plans, allowing us to understand how they differ from what might be reasonably expected.

```{r dem-distrs, layout="l-body-outset", fig.width=8, fig.height=4.5, fig.cap=cap}
cap = str_glue("Democratic vote by district for the staff and preliminary plans and for ",
               scales::number(ncol(as.matrix(subset_sampled(plans$tol_01))), big.mark=","),
               " alternative plans. ",
               "[Learn more about this plot](../../methods.html#district-vote-share-plots)")
rename(plans$tol_01, dem_16=dem) %>%
    plot_dem_distr() +
    labs(title="Average election 2016–2020")
```

Compared to the preliminary plan, the staff plan makes 1 Republican district and 1 Democratic district more competitive.
The nearly-evenly-split district (numbered 4 above) remains highly competitive.


# Proportionality and Representativeness

All our plans are given two R.A. Plan Scores, one for proportionality---how close the seats distribution reflects the statewide vote---and one for representativeness---the odds that an average voter is represented by their preferred political party.
The scores for Colorado are [shown above](#plans-scores).
The graph below shows the proportionality and representativeness for all of our simulated redistricting plans, and the two proposed plans.

```{r prop-repr, fig.width=7, fig.height=4, fig.cap="Proportionality and representativeness for the redistricting plans. [Learn more about these measures](../../methods.html#our-scoring-system)"}
plot_prop_repr(plans$tol_01)
```

Neither plan is very proportional, giving Republicans slightly more seats, on average, than would be otherwise expected.
But the preliminary plan was more representative, with fewer competitive districts.
This meant that more voters were expected to be represented by someone of their party.

What would a good plan look like according to these two measures?
The map below is the simulated redistricting plan which scored the combined best on proportionality and representativeness.
It has two safe Republican seats, one lean-Republican seat, four safe Democratic seats, and one competitive seat.

```{r best, fig.width=7, fig.height=4.5, fig.cap="The sampled plan with the best combination of proportionality and representativeness, with Democratic vote shares labelled."}
best = plans$tol_01 %>%
    filter(abs(proportion) < 1e-4) %>%
    #mutate(score = scale(-abs(proportion))[,1] + scale(represent)[,1]) %>%
    arrange(desc(represent))
best = plans$tol_01 %>%
    filter(draw == as.character(best$draw[1]))
distrs = st_as_sf(as_tibble(map)) %>%
    mutate(district = as.integer(last_plan(best))) %>%
    group_by(district) %>%
    summarize() %>%
    left_join(select(best, district, dem), by="district")
plot_cds(map, last_plan(best), county, "CO") +
    geom_sf_label(aes(label=scales::percent(dem, 1)), data=distrs, fontface="bold",
                  label.padding=unit(0.15, "lines"), label.r=unit(0.05, "lines"),
                  fill="#ffffffaa", label.size=0, inherit.aes=F)
```
